Incremental tactics

### 2012-05-25 doc/papers/LFMTP2012/paper.tex
paper: intro wip

When writing critical software, high assurance against bugs is an
absolute necessity. To certify the absence of bugs in a large body of
code given its specification, human reviewing is one option, but
certainly a failible one. Formal, mathematical proof that our program
respects a certain \emph{dependability claim} gives us the assurance
that ``its trustworthyness [\ldots] would presumably last for eternity
\cite{shao2010certified}''.

Certifying

\cite{miller2011proposal}

%% Describe the problem (Use an example)
%% State your contributions (Bulleted list of contributions)

@@

A student is learning about System \sysname T, and wants to write a
program computing $(2+2)\times 3$. He (or she) knows that sometimes those
$\cst{rec}$ arguments are quite hard to get correct, but System
\sysname T is a typed language; he wants to take advantage of this to
build his program incrementally using the teacher's provided
incremental verifier. As a first attempt, he wants to verify $P_1$:
$$
\recb{\s\z}{\s\z}{\lam x\lam y \s {\var x}}\ \text{.}
$$
For this, he sends the command $\infer{P_1}$ to the system. The
$\infer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function $\lam x\lam y\s{\var
  x}$, and he could reuse the pieces of derivation just built. Let us
call them $\md_2$, $\md_3$ and $\md_4$. Provided there is an operator
$\get\md$ from the derivation to the program it types, what needs really
to be sent to the system is:
\begin{align*}
  &\infer{(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} {\get{\md_4}}}
    \app{\var{add}}\app{\get{\md_2}}{\get{\md_2}}
  )}
\ \text{,}
\end{align*}
that is, only the subterms of the program that changed. Derivations
$\md_2$ and $\md_4$ used to live in the empty context, we are then
free to use it in the new context where
$\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T enjoys
the \emph{weakening} property. Taking advantage of $\md_2$ and $\md_4$
the system succeeds with a new derivation made out of these, without
having to regenerate them because it knows that $\infer{\get\md} =
\md$. Let us call $\md_5$ the subderivation generated for $\tlam x
{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y}
{\get{\md_4}}$, $\md_6$ for the arguments
$\app{\get{\md_2}}{\get{\md_2}}$ and $\md_7$ for the call
$\app{\var{add}} {\get{\md_6}}$. Even if it occurs originally in the
scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
want to reuse it, we will have to provide a proof of this fact.

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \infer{}(
  &\letb{add}{\get{\md_5}}
  \\
  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\get{\gsubst{\md_7}{\msubst H {\infer{\var{add}}}}})}
  {(\s{\get{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\infer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so just use it.

Finally, he decides (in an odd move) to inline the definition of
$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
definition of $\var{mul}$ (it depends on an $\var H$ of
$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
$\times$. He orders:
\begin{align*}
\infer{}(
\letb{add}{\get{\md_5}}
\get{({\gsubst{\md_9} {\msubst {H} {(\get{\var{add}})}; \msubst {H'}
      {\gsubst{\md_8}{\msubst H {(\get{\var{add}})}}}}})}
)
  \ \text{.}
\end{align*}
The definition for $\var{add}$ is unchanged, but the instantiation of
open derivations enables to directly substitute one in the other and
get the appropriate program: call $\md_{10}$ the whole derivation just
returned, $\get{\md_{10}}$ returns:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\&\quad
  {\tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}}}
  \\&\quad\quad
  \app{(\app {\var{add}} \app {(\s{\s\z})}
    {(\s{\s\z})})}{(\s{\s{\s\z}})}
  \ \text{.}
\end{align*}

@@

A \emph{signature} declares object and type constants to be used in
objects. To each object constant we assign an annotation $L$ setting
the slicing behaviour of the constant during type-checking: either
sliceable $\annsliceable$ (each application of this constant will be
cached and given a name), non-sliceable $\annnonsliceable$ (we don't
create a name for applications of this constant), or defined
$\anndefined{T}$ (see \ref{sec:computational-calculus}).
\begin{align*}
  L &\gequal
  \annsliceable \gor
  \annnonsliceable \gor
  \anndefined{T}
  \\
  \Sigma &\gequal
  \gnil \gor
  \gcons \Sigma {\cst c :^L A} \gor
  \gcons \Sigma {\cst a : K}
  &\text{Signature}
\end{align*}
We write $\Sigma(\cst c) = A; L$ for $(\cst c :^L A) \in \Sigma$
suggesting the implementation of signatures as maps.

@@

>>>>

%% Describe the problem (Use an example)

The positive impact of rich type systems on programming language
design is not to be demonstrated anymore, both in terms of the safety
of programs --- they ``do not go wrong'', and programming practices
\eg\ modularity, early bug catching. Verifying well-typing of
real-world programs is thus a critical matter, but it is a complex
operation, for the compiler designer and for the human operator.

\paragraph{For the compiler designer}

Type systems are usually presented in a declarative manner as a set of
inference rules, together with proofs that all programs accepting a
derivation in this system respect a certain dependability claim, the
famous ``progress and preservation'' pair. Yet, actual algorithms of
verification often differ very much from their declarative
description. This is of course true when programs are provided with
little type annotations (type inference), but is also true in many
cases for what is usually called ``type checking'', or
``church-style'' calculi. Take for example dependent type systems,
relying on a certain notion of definitional equality on types, or type
systems with subtype polymorphism: both rely on a rule of the form
$$
\infer{\Gamma\vdash M : A \\ A\leq B}{\Gamma\vdash M : B}\ \text{,}
$$
for a given notion of $\leq$, which is not syntax-directed. The
implementer then has to convert the declarative system to an
algorithm, and convince himself that both are equivalent. This task,
often neglected, is nonetheless highly non-trivial and non-modular.

The most common option is to formally prove the equivalence, \ie\ that
the (typing) relation is actually a function. It is of course the most
satisfactory because it allows to ``forget'' about the declarative
system and take only the result of the algorithm into consideration,
although these proofs can be difficult to obtain. But for the sake of
the discussion, let us choose another path: derivations in the
declarative systems, if they cannot always be algorithmically
inferred, can be \emph{verified} if provided by an oracle. This way,
even if the oracle is wrong, we have a way to decide it. Let us then
imagine a type-checker to be the pair of 1/ an untrusted procedure
$\finfer M$ for inferring a typing derivations out of programs $M$; it
doesn't have to be complete, or even sound, and 2/ a trusted kernel
$\fcheck{\md}$ for verifying these derivations in a given set of
inference rules $\Sigma$. This scheme has the following advantages
over the first approach:
\begin{itemize}
\item it is lightweight: the untrusted algorithm is usually shorter
  than the corresponding equivalence proof
\item it is certifying: if $\fcheck{\finfer M} = \cst{true}$ then
  there really \emph{is} a derivation $\md$ for $M$. Of course the
  converse is not necessarily true: function $\finfer{}\!\!$ could
  fail, or have bugs (return an incorrect derivation).
\item it inherits benefits from certifying approaches like
  proof-carrying code \cite{necula1997proof}: %TODO third-party
                                              %verification
\item for optimization purpose, compilers often need to propagate type
  information in the chain of transformation; the typing derivation
  produced by $\finfer{}$ being explicit, it could constitute the data
  structure being exploited.
\item \ldots\ but there is more:
\end{itemize}

\paragraph{For the programmer}

Compilers are generally monolithic programs, taking source code and
producing machine code in a series of transformations, one of the
first being type checking. But as type systems become richer, it
becomes increasingly difficult for the programmer to write a correct
program in one try, compile it and run it: writing a program longer
than a few dozen of lines becomes a tight and non-linear interaction
between the human and the type-checker involving constant experiments,
fixes, etc. If this interaction is possible for short programs and
fast checkers, it becomes increasingly difficult %TODO

%% State your contributions (Bulleted list of contributions)

@@

A student (the first author) is learning about System \sysname T, and
wants to write a program computing $(2+2)\times 3$. He knows that
sometimes those $\cst{rec}$ arguments are quite hard to get correct,
but System \sysname T is a typed language; he wants to take advantage
of this to build his program incrementally using the teacher's (the
first author too) provided incremental verifier. As a first attempt,
he wants to verify $P_1$:
$$
\recb{\s\z}{\s\z}{\lam x\lam y \s {\var x}}\ \text{.}
$$
For this, he sends the command $\finfer{P_1}$ to the system. The
$\finfer{}$ operator takes a well-formed program and tries to construct
a valid typing derivation for its argument in the current, empty
context. The system succeeds and returns a typing derivation $\md_1$
of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.

The student then decides to factorize a definition for $\var{add}$ out
of his code:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\
  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
\end{align*}
But he (or his text editor) realizes that some parts of this program
have not changed: both $\s\z$, and the function $\lam x\lam y\s{\var
  x}$, and he could reuse the pieces of derivation just built. Let us
call them $\md_2$, $\md_3$ and $\md_4$. Provided there is an operator
$\get\md$ from the derivation to the program it types, what needs really
to be sent to the system is:
\begin{align*}
  &\finfer{(
    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
      {\var y} {\get{\md_4}}}
    \app{\var{add}}\app{\get{\md_2}}{\get{\md_2}}
  )}
\ \text{,}
\end{align*}
that is, only the subterms of the program that changed. Derivations
$\md_2$ and $\md_4$ used to live in the empty context, we are then
free to use it in the new context where
$\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T enjoys
the \emph{weakening} property. Taking advantage of $\md_2$ and $\md_4$
the system succeeds with a new derivation made out of these, without
having to regenerate them because it knows that $\finfer{\get\md} =
\md$. Let us call $\md_5$ the subderivation generated for $\tlam x
{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y}
{\get{\md_4}}$, $\md_6$ for the arguments
$\app{\get{\md_2}}{\get{\md_2}}$ and $\md_7$ for the call
$\app{\var{add}} {\get{\md_6}}$. Even if it occurs originally in the
scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
want to reuse it, we will have to provide a proof of this fact.

The student then provides a definition for multiplication, and writes
the operation $(2 + 2)\times3$:
\begin{align*}
  \finfer{}(
  &\letb{add}{\get{\md_5}}
  \\
  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}} }
  \\
  &\quad\quad\app{\var{mul}}\app
  {(\get{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
  {(\s{\get{\md_2}})}
  )
  \ \text{.}
\end{align*}
Here again, he took care of reusing all derivations for equal
subterms. In particular, $\md_7$ expects a proof that the variable
$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
$\finfer{\var{add}}$ meaning: at this point we will have generated (or
reused in this case) the derivation of $\var{add}$, so just use it.

Finally, he decides (in an odd move) to inline the definition of
$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
definition of $\var{mul}$ (it depends on an $\var H$ of
$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
$\times$. He orders:
\begin{align*}
\finfer{}(
\letb{add}{\get{\md_5}}
\get{({\gsubst{\md_9} {\msubst {H} {(\get{\var{add}})}; \msubst {H'}
      {\gsubst{\md_8}{\msubst H {(\get{\var{add}})}}}}})}
)
  \ \text{.}
\end{align*}
The definition for $\var{add}$ is unchanged, but the instantiation of
open derivations enables to directly substitute one in the other and
get the appropriate program: call $\md_{10}$ the whole derivation just
returned, $\get{\md_{10}}$ returns:
\begin{align*}
  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
    {\var y} {\lam x \lam y \s{\var x}}}
  \\&\quad
  {\tlam x {\cst{nat}} \tlam y
    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
      \app{\var{add}}\app{\var x} {\var z}}}
  \\&\quad\quad
  \app{(\app {\var{add}} \app {(\s{\s\z})}
    {(\s{\s\z})})}{(\s{\s{\s\z}})}
  \ \text{.}
\end{align*}

@@

A \emph{signature} declares object and type constants to be used in
objects. To each object constant we assign an annotation $L$ setting
the slicing behaviour of the constant during type checking: either
sliceable $\annsliceable$ (each application of this constant will be
cached and given a name), non-sliceable $\annnonsliceable$ (we don't
create a name for applications of this constant), or defined
$\anndefined{T}$ (see \ref{sec:computational-calculus}).
\begin{align*}
  L &\gequal
  \annsliceable \gor
  \annnonsliceable \gor
  \anndefined{T}
  \\
  \Sigma &\gequal
  \gnil \gor
  \gcons \Sigma {\cst c :^L A} \gor
  \gcons \Sigma {\cst a : K}
  &\text{Signature}
\end{align*}
We write $\Sigma(\cst c) = A; L$ for $(\cst c :^L A) \in \Sigma$
suggesting the implementation of signatures as maps.

@@

%  LocalWords:  subtype
